Causal Inference in Population Trends: Searching for Demographic Anomalies in Big Data

### 2020-09-14 MainDocument/MainDocument.Rmd
Update MainDocument.Rmd

# abstract: "The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms can uncover hidden anomalies in demographic data that have previously been dismissed as noise. Here, we demonstrate how social scientists can use causal inference techniques and abductive reasoning to identify fertility and mortality anomalies in state-level data trends by investigating demographic time series data since 1999. In demonstrating our technique, we uncover real spikes and dips in fertility and mortality over time, distinguished from regular trend variations. We show that 36 states exhibited at least one mortality anomaly over time, totaling nearly 300,000 additional deaths, while an additional seven states exhibited periods of decline resulting in 250,000 fewer deaths. Ten states exhibited real, marked periods of declines in fertility, totaling more than 214,000 fewer births than expected, and 8 states had periods of excess births, resulting in an addition 12,000 births. Our results show how researchers can detect demographic anomalies in trend data and use those anomalies as a beginning point for further investigation into the causes."

@@

The search for causality is one of the primary motivations behind social science [@smith2011commentary; @pearl2009causal]. Causal Inference, the identification of causal mechanisms, attempts to uncover the underlying mechanisms that result in changes in a phenomenon and has a long history as an approach to inquiry in the social sciences [@grimmer2015ppsp]. Often, the identification of a casual mechanism is accomplished via a randomized control trial (RCT) or expert knowledge applied to a natural experiment [@salganik2019bit].

Unfortunately, both RCTs and natural experiments have drawbacks. RCTs are often prohibitively expensive for many scientists, precluding their widespread adoption in the social sciences [@west2008ajph]. RCTs are also sometimes impossible to deploy in the "real world". For instance, an RCT to study country-level austerity measures would be difficult, if not impossible, to implement. Natural experiments are a cheaper alternative, but provide less certainty about causation, fewer opportunities to replicate work, and require expert knowledge of the natural experiment to be properly employed [@pearl2018book].

@@

Additionally, both RCTs and natural experiments follow traditional hypothesis testing, the deductive scientific paradigm where a question is first posed and scientists generate or find data to answer the question. Such approaches, while extensively utilized and time-tested, are likely to miss important phenomena that might go unnoticed or suffer from extensive p-hacking [@head2015extent;@nuzzo2014scientific;@ruggles2014big]. Abductive reasoning, sometimes described as “inferring cause from effect” [@Crowder2017], moves the scientific process from the inductive to the deductive, and sometimes back and forth, to reach conclusions [@bryant2014realm] The physical sciences use abductive reasoning more fully compared to the social sciences. The discovery of Neptune in 1846 based on unexplained perturbations of the orbit of Uranus [@popper2005logic], the discovery of galactic migration of planets [@gomes2005n], and the current generation of gravitational wave detectors such as the Laser Interferometer Gravitational-Wave Observatory (LIGO) [@harry2010advanced] demonstrate the power of the abductive approach in the physical sciences.

In a data-rich environment, abductive reasoning can be more easily applied to first find effects and then infer their cause ex-post. The modern proliferation of data, advances in high performance (“super”) computing, and the development of powerful statistical algorithms provide an environment ripe for wedding causal inference with abductive reasoning [@van2016data;@zikopoulos2011], with the potential for researchers to find hidden or understudied social phenomena [@bohon2018demography]. While the availability of Big Data and high performance computing allows novel exploration of data through causal inference [@bohon2018demography;@rcausalimpact;@shiffrin2016drawing], relatively few studies utilize causal inference techniques in the study of demographic phenomena. However, understanding social phenomena using these advances reveals important insights into society [@angrist1989lifetime;@mas2009peers] and allows us to better monitor population trends [@nobles2019;@torche2015hidden]. The rich demographic data available in the United States make the potential revelation of interesting and important demographic phenomena not only possible, but extremely plausible – even if identification of the phenomena occurs without identifying the underlying cause.

@@

In this paper, we combine a modern statistical outlier detection algorithm [@chen1993joint], abductive reasoning, and large demographic data sets on more than 20 years of mortality/fertility to identify anomalous demographic behavior. In essence, we identify *effects without knowledge of the cause*. We ask one fundamental question regarding demographic anomalies: What are the hidden baby booms/busts and mortality spikes/dips in the United States over the last twenty years? We do not necessarily know the causes of these anomalies but identifying them allows scientists with more detailed knowledge of local population dynamics, state-level policy making, or macro-economics to explain these phenomena post-hoc. From explanations of phenomena that may have previously gone unnoticed, social scientists may be able to better forecast populations and provide policy solutions for impending problems such as climate change.

@@

Scientific inquiry falls under one of three modes of logical reasoning: deduction, induction, and abduction. Deductive reasoning starts with a theory and makes a prediction of what the observations should be if that theory is correct. The classic example of deductive reasoning essentially goes (a) "all men are mortal", (b) "Socrates is a man", (c) "Therefore, Socrates is mortal." Deductive reasoning has a long history in the sciences and most hypothesis testing in the social sciences follows this logical reasoning. For example, John Snow used deductive reasoning in to stop the 1854 London cholera outbreak. He hypothesized that cholera was waterborne and deduced that a bad water pump was responsible for the outbreak.

Inductive reasoning is the opposite of deductive reasoning. Rather than starting with a theory and making a prediction, induction starts with an observation and makes a theory based on the likeliest reason. For example, the Hispanic Mortality Paradox first appeared in the scientific literature in the mid-1980s [@markides1986health] as an observation that hispanics in the United States had, paradoxically, better health outcomes than expected given their socioeconomic status. Based on this observation, numerous theories arose to explain this paradox.

Both deductive and inductive reasoning are well established in science but are not without their shortcomings. In both approaches, it is impossible, or at least imporbable, for the premise to be true and the conclusion false. Furthermore, if a hypothesis is rejected under a deductive or inductive approach, there is no alternative explanation -- there is either the presence or absence of support.

@@

Abductive reasoning and exploratory data analysis are best employed in a data rich environment -- the very data environment social scientists and demographers find themselves in the present moment. Additionally, in a data rich environment, a hypothesis is a hidden liability [@yanai_hypothesis_2020]because it focuses our attention on only that which we seek^[Yanai and Lercher [-@yanai_hypothesis_2020] ask if focusing on a specific hypothesis prevents new discovery? They created a simple dataset that if plotted would reveal an image of a gorilla waving. Students instructed to find support for specific hypotheses were less likely to discover the gorilla compared to those who were simply asked "What do you conclude from the dataset?"]. Focusing on specific hypotheses often prevents us from finding other interesting phenomena. Afterall, it was the publication of the hispanic paradox itself that has generated more than thirty years of devoted scholarship. When paired with causal inference techniques, abductive reasoning provides an empirical platform for the discovery of potentially interesting social phenomena; it provides an important tool for sifting large amounts of data to find interesting results in the absence of hypothesis testing.

@@

    df2<- tso(y, cval = 3.5)$outliers
    df2$iteration <- i
    results_AO <- bind_rows(results_AO, df2)
    # print(i)
  } , error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}
```
In our anomaly-less time series, we detect `r nrow(results_noutliers)` anomalies on the 180,000 candidate time steps for a false-positive error rate of `r prettyNum(nrow(results_noutliers) / 180000*100, digits = 3)`%. In our time series containing a single anomaly, we detect `r nrow(results_AO[which(results_AO$time == 150),])` anomalies out of the 1000 candidate time series, for a false-negative error rate of `r (1-(nrow(results_AO[which(results_AO$time == 150),]) / 1000)) * 100`%. This provides a statistical power of `r nrow(results_AO[which(results_AO$time == 150),]) / 1000`. These results give us confidence that the anomalies we detect in this paper are true anomalies.

@@

In **\autoref{fig:mortnewhamp}a**, notice the strong level shift (LS) that occurs in February 2004 (2004:02 *t*= `r paste(ny_tstat)`) which prevented `r paste(ny_averteddeaths)` deaths per month. This shift totals more than `r paste(ny_averteddeathstot)` averted deaths compared to the counter-factual time series and is the single largest mortality protective anomaly among all states. This translates to `r ny_peraverted` fewer deaths than expected over the time period. What is driving this mortality protection? What policies did NY put into place that might have contributed to this considerable mortality reduction? What environmental or economic conditions may have changed may have changed? These are the kinds of questions that arise from our analyses. The purpose of our paper is not to answer these questions, but our findings underscore the need for more research that takes an abductive approach. By identifying anomalies through an inductive process, researchers can then look for underlying causes. Once those causes are identified, researchers can then use a deductive process to see if such events predict other (or future) anomalies.

@@

This is contrast to **\autoref{fig:ferthawaii}b**, where we identify a strange mortality reduction in Ohio (*t*-stat: `r oh_tstat`). This LS is more than `r oh_deaths` deaths per month less than the counter-factual time series, suggesting had nearly `r oh_tot` fewer deaths since February 2015 than expected. This is the single-largest LS among all states. We could not identify the potential policies Ohio might have put into place to provide such a strong mortality protection.

@@

Consistent with more anomalies across the time series, we find more anomalous deaths (`r paste(prettyNum(sums[2,6], digits = 0, big.mark = ","))`) than anomalous births (`r paste(prettyNum(sums[1,6], digits = 0, big.mark = ","))`). Fertility anomalies overwhelmingly tend toward lower fertility, with only a few fertility anomalies yielding more births. This is consistent with research findings that link high severity weather events to significantly lower fertility [@Evans2008Hurricanebirth]. Conversely, mortality anomalies tend to be more evenly split between positive and negative anomalies, but tend toward more deaths rather than fewer deaths. Again, this is not surprising, as the kinds of events that increase death (e.g., plant closings are associated with increased opioid death [@Venkataramani2019]) are more common than the events that decrease death (e.g., cannabis legalization is associated with decreased opioid deaths [@Livingston2017Cannabis]). These anomalous deaths and births account for `r sums2$percent_anom[2]`% and `r sums2$percent_anom[1]`% over the entire time series, respectively.

As **\autoref{fig:summap}** shows, three states exhibited neither mortality nor fertility anomalies: Alaska, North Dakota, and South Dakota. Another nine states exhibited just a single anomaly (DE, DC, ID, IL, KS, NV, NM, UT, and WY). Both New York and Massachusetts exhibited the most anomalies with nine.

@@

# Conclusion
Data scientists frequently claim that the big data revolution is a turning point in scientific discovery that will allow us to solve some of the world’s most pressing problems [@grimmer2015ppsp].  Social scientists are skeptical of such claims, because they better understand the complexities of the social world and know from experience that data alone is not enough to solve social problems [@bohon2018demography;@grimmer2015ppsp]. Nonetheless, the increased availability of data and (more importantly, we argue) the development of advanced techniques for analyzing these data will enable important discovery [@monroe2015no].

>>>>

# abstract: "The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms allows us to uncover hidden anomalies in demographic data that have previously been dismissed as noise. Here, we demonstrate how social scientists can use causal inference techniques and abductive reasoning to identify fertility and mortality anomalies in state-level data trends by investigating demographic time series data since 1999. In demonstrating our technique, we uncover real spikes and dips in fertility and mortality over time, distinguished from regular trend variations. We show that 36 states exhibited at least one mortality anomaly over time, totaling nearly 300,000 additional deaths, while an additional seven states exhibited periods of decline resulting in 250,000 fewer deaths. Ten states exhibited real, marked periods of declines in fertility, totaling more than 214,000 fewer births than expected, and 8 states had periods of excess births, resulting in an addition 12,000 births. Our results show how researchers can detect demographic anomalies in trend data and use those anomalies as a beginning point for further investigation into the causes."

@@

The search for causality is one of the primary motivations of social scientists [@smith2011commentary; @pearl2009causal]. Causal inference, the identification of causal mechanisms, attempts to uncover the underlying mechanisms that result in changes in a phenomenon and has a long history as an approach to inquiry in the social sciences [@grimmer2015ppsp]. Often, the identification of a casual mechanism is accomplished via a randomized control trial (RCT) or expert knowledge applied to a natural experiment [@salganik2019bit].

Unfortunately, both RCTs and natural experiments have drawbacks. RCTs are often prohibitively expensive for many scientists, precluding their widespread adoption in the social sciences [@west2008ajph]. RCTs are also sometimes impossible to deploy in the "real world." For instance, an RCT to study country-level austerity measures would be difficult, if not impossible, to implement. RCTs to study fertility would be immoral. Natural experiments are a cheaper alternative but provide less certainty about causation, fewer opportunities to replicate work, and require expert knowledge of the natural experiment to be properly employed [@pearl2018book].

@@

Additionally, both RCTs and natural experiments are based on traditional hypothesis-based testing, the deductive scientific paradigm where a question is first posed and scientists generate or find data to answer the question. Such approaches, while extensively utilized and time-tested, increase the likelihood that a researcher will miss important phenomena that might go unnoticed because it falls outside the hypothesis, or the study could suffer from extensive p-hacking [@head2015extent;@nuzzo2014scientific;@ruggles2014big]. At the same time, purely inductive data mining can result in false assertions or spurious findings, especially in the hands of those who are not field experts [@yanai_hypothesis_2020]. Abductive reasoning, sometimes described as “inferring cause from effect” [@Crowder2017], moves the scientific process from the inductive to the deductive, and sometimes back and forth, to reach conclusions [@bryant2014realm] Physical scientists use abductive reasoning more commonly than social scientists. The discovery of Neptune in 1846 based on unexplained perturbations of the orbit of Uranus [@popper2005logic], the discovery of galactic migration of planets [@gomes2005n], and the current generation of gravitational wave detectors such as the Laser Interferometer Gravitational-Wave Observatory (LIGO) [@harry2010advanced] demonstrate the power of the abductive approach in the physical sciences.

In a data-rich environment, abductive reasoning can be more easily applied to first find effects and then infer their cause ex-post. The modern proliferation of data, advances in high performance (“super”) computing, and the development of powerful statistical algorithms provide an environment ripe for wedding causal inference with abductive reasoning [@van2016data;@zikopoulos2011], with the potential for researchers to find hidden or understudied social phenomena [@bohon2018demography]. While the availability of big data and high performance computing allows novel exploration of data through causal inference [@bohon2018demography;@rcausalimpact;@shiffrin2016drawing], relatively few studies utilize causal inference techniques in the study of demographic phenomena. However, understanding social phenomena using these advances reveals important insights into society [@angrist1989lifetime;@mas2009peers] and allows us to better monitor population trends [@nobles2019;@torche2015hidden]. The rich demographic data available in the United States make the potential revelation of interesting and important demographic phenomena not only possible, but extremely plausible – even if identification of the phenomena occurs without identifying the underlying cause.

@@

In this paper, we combine a modern statistical outlier detection algorithm [@chen1993joint], abductive reasoning, and large demographic data sets on more than 20 years of mortality and fertility trends to identify anomalous demographic behavior. In essence, we identify *effects without knowledge of the cause*. We ask one fundamental question regarding demographic anomalies: What are the hidden baby booms/busts and mortality spikes/dips in the United States over the last twenty years? We do not necessarily know the causes of these anomalies but identifying them allows scientists with more detailed knowledge of local population dynamics, state-level policy making, or macro-economics to explain these phenomena post-hoc. From explanations of phenomena that may have previously gone unnoticed, social scientists may be able to better forecast populations and provide policy solutions for impending problems such as climate change.

@@

Scientific inquiry falls under one of three modes of logical reasoning: deduction, induction, and abduction. Deductive reasoning starts with a theory and makes a prediction of what the observations should be if that theory is correct. The classic example of deductive reasoning essentially goes (a) "all men are mortal," (b) "Socrates is a man," (c) "therefore, Socrates is mortal." Deductive reasoning has a long history in the sciences and most hypothesis testing in the social sciences follows this logical reasoning. For example, John Snow used deductive reasoning to stop the 1854 London cholera outbreak. He hypothesized that cholera was waterborne and deduced that a bad water pump was responsible for the outbreak.

Inductive reasoning is directionally opposite of deductive reasoning. Rather than starting with a theory and making a prediction, induction starts with an observation and makes a theory based on the likeliest reason. For example, the Hispanic Mortality Paradox first appeared in the scientific literature in the mid-1980s [@markides1986health] as an observation that Hispanics in the United States had better health outcomes than expected given their socioeconomic status. Based on this observation, numerous theories arose to explain this paradox.

Both deductive and inductive reasoning are well established in science but are not without their shortcomings. In both approaches, it is impossible, or at least improbable, for the premise to be true and the conclusion false. Furthermore, if a hypothesis is rejected under a deductive or inductive approach, there is no alternative explanation -- there is either the presence or absence of support.

@@

Abductive reasoning and exploratory data analysis are best employed in a data rich environment -- the very data environment social scientists and demographers find themselves in the present moment. In a data rich environment, a hypothesis is a hidden liability [@yanai_hypothesis_2020] because it focuses our attention on only that which we seek. To demonstrate this, ^[Yanai and Lercher [-@yanai_hypothesis_2020] ask if focusing on a specific hypothesis prevents new discovery. By way of illustration, they created a simple dataset that if plotted would reveal an image of a gorilla waving. Students instructed to find support for specific hypotheses were less likely to discover the gorilla compared to those who were simply asked "What do you conclude from the dataset?"]. Focusing on specific hypotheses often prevents us from finding other interesting phenomena. After all, it was the publication of the Hispanic paradox itself that has generated more than thirty years of devoted scholarship. At the same time, when paired with causal inference techniques, abductive reasoning provides an empirical platform for the discovery of potentially interesting social phenomena; it provides an important tool for sifting large amounts of data to find interesting results without the blinders of hypothesis testing.

@@

    df2<- tso(y, cval = 3.5)$outliers
    df2$iteration <- i
    results_AO <- bind_rows(results_AO, df2)
    # print(i)
  } , error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}
```
In our sans-anomaly time series, we detect `r nrow(results_noutliers)` anomalies on the 180,000 candidate time steps for a false-positive error rate of `r prettyNum(nrow(results_noutliers) / 180000*100, digits = 3)`%. In our time series containing a single anomaly, we detect `r nrow(results_AO[which(results_AO$time == 150),])` anomalies out of the 1000 candidate time series, for a false-negative error rate of `r (1-(nrow(results_AO[which(results_AO$time == 150),]) / 1000)) * 100`%. This provides a statistical power of `r nrow(results_AO[which(results_AO$time == 150),]) / 1000`. These results give us confidence that the anomalies we detect in this paper are true anomalies.

@@

In **\autoref{fig:mortnewhamp}a**, notice the strong level shift (LS) that occurs in February 2004 (2004:02 *t*= `r paste(ny_tstat)`) which prevented `r paste(ny_averteddeaths)` deaths per month. This shift totals more than `r paste(ny_averteddeathstot)` averted deaths compared to the counter-factual time series and is the single largest mortality protective anomaly among all states. This translates to `r ny_peraverted` fewer deaths than expected over the time period. What is driving this mortality protection? What policies did NY put into place that might have contributed to this considerable mortality reduction? What environmental or economic conditions may have changed? These are the kinds of questions that arise from our analyses. The purpose of our paper is not to answer these questions, but our findings underscore the need for more research that takes an abductive approach. By identifying anomalies through an inductive process, researchers can then look for underlying causes. Once those causes are identified, researchers can then use a deductive process to see if such events predict other (or future) anomalies.

@@

This is contrast to **\autoref{fig:ferthawaii}b**, where we identify a strange mortality reduction in Ohio (*t*-stat: `r oh_tstat`). This LS is more than `r oh_deaths` deaths per month less than the counter-factual time series, suggesting had nearly `r oh_tot` fewer deaths since February 2015 than expected. This is the single-largest LS among all states. We could not identify the potential policies Ohio might have put into place or the events that occurred to provide such a strong mortality protection.

@@

Consistent with more anomalies across the time series, we find more anomalous deaths (`r paste(prettyNum(sums[2,6], digits = 0, big.mark = ","))`) than anomalous births (`r paste(prettyNum(sums[1,6], digits = 0, big.mark = ","))`). Fertility anomalies overwhelmingly tend toward lower fertility, with only a few fertility anomalies yielding more births. This is consistent with research findings that link severe weather events to significantly lower fertility [@Evans2008Hurricanebirth]. Conversely, mortality anomalies tend to be more evenly split between positive and negative anomalies, but tend toward more deaths rather than fewer deaths. Again, this is not surprising, as the kinds of events that increase death (e.g., plant closings are associated with increased opioid death [@Venkataramani2019]) are more common than the events that decrease death (e.g., cannabis legalization is associated with decreased opioid deaths [@Livingston2017Cannabis]). These anomalous deaths and births account for `r sums2$percent_anom[2]`% and `r sums2$percent_anom[1]`% over the entire time series, respectively.

As **\autoref{fig:summap}** shows, three states exhibited neither mortality nor fertility anomalies: Alaska, North Dakota, and South Dakota. Another nine states exhibited just a single anomaly (Delaware, District of Colombia, Idaho, Illinois, Kansas, Nevada, New Mexico, Utah, and Wyoming). Both New York and Massachusetts exhibited the most anomalies with nine.

@@

# Conclusion
Data scientists frequently claim that the big data revolution is a turning point in scientific discovery that will allow us to solve some of the world’s most pressing problems [@grimmer2015ppsp].  Social scientists are skeptical of such claims, because they better understand the complexities of the social world and know from experience that data alone are not enough to solve social problems [@bohon2018demography;@grimmer2015ppsp]. Nonetheless, the increased availability of data and (more importantly, we argue) the development of advanced techniques for analyzing these data will enable important discovery [@monroe2015no].
